\chapter{Contents of the attached media}


\dirtree{%
.1 .changeset.
.2 config.json.
.1 .eslintrc.
.1 .github.
.2 workflows.
.3 main.yml.
.3 publish.yml.
.1 .gitignore.
.1 .npmignore.
.1 .prettierrc.
.1 .vscode.
.2 settings.json.
.1 assets.
.2 cover.svg.
.1 CHANGELOG.md.
.1 package.json.
.1 README.md.
.1 scripts.
.2 dirtree.ts.
.2 generate.ts.
.2 parser.ts.
.1 src.
.2 expression.
.3 enum.ts.
.3 evaluator.test.ts.
.3 evaluator.ts.
.3 lexer.test.ts.
.3 lexer.ts.
.3 parser.test.ts.
.3 parser.ts.
.2 hotscript.test.ts.
.2 hotscript.ts.
.2 index.ts.
.2 math.
.3 abs.test.ts.
.3 abs.ts.
.3 add.test.ts.
.3 add.ts.
.3 ceil.test.ts.
.3 ceil.ts.
.3 comparison.test.ts.
.3 comparison.ts.
.3 divide.test.ts.
.3 divide.ts.
.3 factorial.test.ts.
.3 factorial.ts.
.3 floor.test.ts.
.3 floor.ts.
.3 modulo.test.ts.
.3 modulo.ts.
.3 multiply.test.ts.
.3 multiply.ts.
.3 negate.test.ts.
.3 negate.ts.
.3 power.test.ts.
.3 power.ts.
.3 root.test.ts.
.3 root.ts.
.3 round.test.ts.
.3 round.ts.
.3 subtract.test.ts.
.3 subtract.ts.
.3 truncate.test.ts.
.3 truncate.ts.
.2 utils.
.3 array.ts.
.3 bits.ts.
.3 boolean.ts.
.3 map.ts.
.3 parse.test.ts.
.3 parse.ts.
.1 thesis.
.2 .gitignore.
.2 .gitlab-ci.yml.
.2 assignment-include.pdf.
.2 ctufit-thesis.acn.
.2 ctufit-thesis.cls.
.2 ctufit-thesis.glo.
.2 ctufit-thesis.glsdefs.
.2 ctufit-thesis.ist.
.2 ctufit-thesis.pdf.
.2 ctufit-thesis.tex.
.2 indentconfig.yml.
.2 LICENSE.
.2 README.md.
.2 sources.bib.
.2 text.
.3 analysis.
.3 appendix.tex.
.3 conclusion.
.4 conclusion.tex.
.3 implementation.
.3 introduction.
.4 introduction.tex.
.3 medium.tex.
.3 testing.
.4 testing.tex.
.1 tsconfig.json.
.1 yarn.lock.
}
